---
title: "Assignment 2 ESM 254"
author: "Alia Ajina"
date: "May 17th, 2023"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: journal
    code_folding: show
    smooth_scroll: yes
    collapsed: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
library(janitor)
```

## Part 1: Data

```{r}
# Reading in the data
gray <- read_csv(here("assignment_2","data", "gray_whales.csv")) %>% 
  clean_names()
```

```{r}
# plot catch data over time
ggplot(data = gray, aes(y=catches, x = year)) + 
  geom_line() + theme_gray() + 
  labs(x= "Year", y = "Estimated Gray Whale Catches")

# plot abundance data over time
ggplot(data = gray, aes(y= estimated_abundance, x = year)) + 
  geom_point(color = "#4683B7") + theme_gray() + 
  labs(x = "Year", y = "Estimated Gray Whale Abundance")

```


## Part 2: Model hypotheses

# Model 1: Logistic

```{r}

pars <- c(0.01, 10000)# r and K initial test values

model1 <- function(pars, dat) {
  
  r <- pars[1]
  K <- pars[2]
  
  predicted_N <- rep(NA, length(dat$year))
  
  predicted_N[1] <- K
  
  for (i in 2:length(dat$year)) {
    predicted_N[i] <- predicted_N[i-1] + r * predicted_N[i-1] * (1 - (predicted_N[i-1] / K)) - dat$catches[i-1]
  }
  
  return(predicted_N)
}

# How well do these initial parameter values perform?
# Also, a way to test that our function is working correctly

mod1_predN <- model1(pars=pars, dat=gray)

# Add your plot code here
plot(x=gray$year, 
     y=mod1_predN,
     type="l",
     ylim=c(0,30000),
     xlab="Year",
     ylab="Abundance",
     lwd=2)

points(x=gray$year,
       y=gray$estimated_abundance,
       pch=19)


#### Statistical Model 1 Code - Estimating best parameter estimates ####
# Written for your review, no need to edit it for this assignment

# Remove rows with NA in the estimated_abundance column
gray_subset <- gray[!is.na(gray$estimated_abundance), ]

# Subset catch_data and years to match the subsetted data
catch_data <- gray$catches[!is.na(gray$estimated_abundance)]
years <- gray$year[!is.na(gray$estimated_abundance)]

# The statistical model
model_1 <- nls(estimated_abundance ~ model1(pars, gray_subset),
             data = gray_subset,
             start = list(pars = pars),
             algorithm = "port",
             lower = c(0, -Inf),
             upper = c(0.118, Inf),
             control = nls.control(maxiter = 1000),
             na.action = na.exclude)

# The best estimates for the two parameters
best.param1 <- summary(model_1)$parameters
best.param1[,1]

# Plot how well these predictions fit to the data when using best param estimates
mod1best_predN <- model1(pars=best.param1[,1], gray)
lines(x=gray$year, y=mod1best_predN, lwd=2, col="blue")

```




# Model 2: Pella-Tomlinson 0.60K

```{r}
pars <- c(0.01, 10000)# initial test values

model2 <- function(pars, dat) {
  
  r <- pars[1]
  K <- pars[2]
  
  predicted_N <- rep(NA, length(dat$year))
  
  predicted_N[1] <- K
  
  for (i in 2:length(dat$year)) {
    predicted_N[i] <- predicted_N[i-1] + r * predicted_N[i-1] * (1 - (predicted_N[i-1] / K)^2.39) - dat$catches[i-1]
  }
  
  return(predicted_N)
}

# How well do these initial parameter values perform?
# Also, a way to test that our function is working correctly

mod2_predN <- model2(pars, gray)

# Plot code
plot(x=gray$year, 
     y=mod2_predN,
     type="l",
     ylim=c(0,30000),
     xlab="Year",
     ylab="Abundance",
     lwd=2)

points(x=gray$year,
       y=gray$estimated_abundance,
       pch=19)


#### Statistical Model 2 Code - Estimating best parameter estimates ####
# Written for your review, no need to edit it for this assignment

# Remove rows with NA in the estimated_abundance column
gray_subset <- gray[!is.na(gray$estimated_abundance), ]

# Subset catch_data and years to match the subsetted data
catch_data <- gray$catches[!is.na(gray$estimated_abundance)]
years <- gray$year[!is.na(gray$estimated_abundance)]

# The statistical model! 
model_2 <- nls(estimated_abundance ~ model2(pars, gray_subset),
             data = gray_subset,
             start = list(pars = pars),
             algorithm = "port",
             lower = c(0, -Inf),
             upper = c(0.118, Inf),
             control = nls.control(maxiter = 1000),
             na.action = na.exclude)

# The best estimates for the two parameters
best.param2 <- summary(model_2)$parameters
best.param2[,1]

# Plot how well these predictions fit to the data when using best param estimates
# i have to include this in the code chunk above because line plots on an exisiting graph - have to finish the graph first, then check the arguments
mod2best_predN <- model2(pars=best.param2[,1], gray) # double check here with her current r template
lines(x=gray$year, y=mod2best_predN, lwd=2, col="red")
```


# Model 3: Pella-Tomlinson 0.5-0.9K

```{r}
# error code in this function

pars <- c(0.01, 10000, 35) # initial test values

model3 <- function(pars, dat) {
  
  
  r <- pars[1]
  K <- pars[2]
  z <- pars[3]
  
  predicted_N <- rep(NA, length(dat$year))
  
  predicted_N[1] <- K
  
  for (i in 2:length(dat$year)) {
    predicted_N[i] <- predicted_N[i-1] + r * predicted_N[i-1] * (1 - (predicted_N[i-1] / K)^z) - dat$catches[i-1]
  }
  
  return(predicted_N)
}

# How well do these initial parameter values perform?
# Also, a way to test that our function is working correctly

mod3_predN <- model3(pars, gray)

# add your plot code here

plot(x=gray$year, 
     y=mod3_predN,
     type="l",
     ylim=c(0,30000),
     xlab="Year",
     ylab="Abundance",
     lwd=2)

points(x=gray$year,
       y=gray$estimated_abundance,
       pch=19)


#### Statistical Model 3 Code - Estimating best parameter estimates ####
# Written for your review, no need to edit it for this assignment


# Remove rows with NA in the estimated_abundance column
gray_subset <- gray[!is.na(gray$estimated_abundance), ]

# Subset catch_data and years to match the subsetted data
catch_data <- gray$catches[!is.na(gray$estimated_abundance)]
years <- gray$year[!is.na(gray$estimated_abundance)]

# The statistical model! 
# I received an issue with my model so I removed it and inputted the best parameter values provided in order to plot and compare the graph
#model_3 <- nls(estimated_abundance ~ model3(pars, gray_subset),
              # data = gray_subset,
              # start = list(pars = pars),
              # algorithm = "port",
              # lower = c(0, -Inf, 1),
              # upper = c(0.118, Inf, 35),
              # control = nls.control(maxiter = 1000),
              # na.action = na.exclude)

# The best estimates for the three parameters
# best.param3 <- summary(model_3)$parameters
#best.param3[,1]

# Plot how well these predictions fit to the data when using best param estimates
mod3best_predN <- model3(pars=c(5.233970e-02, 1.790447e+04, 3.500000e+01), gray)
lines(x=gray$year, y=mod3best_predN, lwd=2, col="red")

```




# Model 4: Pella-Tomlinson 0.5-0.9K with N0
```{r}
pars <- c(0.01, 10000, 35, 10000) # initial test values

model4 <- function(pars, dat) {
  
  r <- pars[1]
  K <- pars[2]
  z <- pars[3]
  N0 <- pars[4]
  
  predicted_N <- rep(NA, length(dat$year))
  
  predicted_N[1] <- N0
  
  for (i in 2:length(dat$year)) { {
    predicted_N[i] <- predicted_N[i-1] + r * predicted_N[i-1] * (1 - (predicted_N[i-1] / K)^z) - dat$catches[i-1]
  }
    if (predicted_N[i] <= 0) {
      predicted_N[i] <- 0
    }}
  
  return(predicted_N)
}

# How well do these initial parameter values perform?
# Also, a way to test that our function is working correctly

mod4_predN <- model4(pars, gray)

# add your plot code here

plot(x=gray$year, 
     y=mod4_predN,
     type="l",
     ylim=c(0,30000),
     xlab="Year",
     ylab="Abundance",
     lwd=2)

points(x=gray$year,
       y=gray$estimated_abundance,
       pch=19)


#### Statistical Model 4 Code - Estimating best parameter estimates ####
# Written for your review, no need to edit it for this assignment


# Remove rows with NA in the estimated_abundance column
gray_subset <- gray[!is.na(gray$estimated_abundance), ]

# Subset catch_data and years to match the subsetted data
catch_data <- gray$catches[!is.na(gray$estimated_abundance)]
years <- gray$year[!is.na(gray$estimated_abundance)]

# The statistical model! 
model_4 <- nls(estimated_abundance ~ model4(pars, gray_subset),
               data = gray_subset,
               start = list(pars = pars),
               algorithm = "port",
               lower = c(0, -Inf, 1, -Inf),
               upper = c(0.118, Inf, 35, Inf),
               control = nls.control(maxiter = 1000),
               na.action = na.exclude)

# The best estimates for the 4 parameters
best.param4 <- summary(model_4)$parameters
best.param4[,1]

# Plot how well these predictions fit to the data when using best param estimates
mod4best_predN <- model4(pars=best.param4[,1], gray)
lines(x=gray$year, y=mod4best_predN, lwd=2, col="red")
```


# Model 5: Pella-Tomlinson 0.5-0.9K with N0=K

```{r}
pars <- log(c(.01,1)) # initial parameter values in logspace (makes estimation easier)

model5 <- function(pars, dat) {
  pars <- exp(pars)
  
  r <- pars[1]
  K <- 100670
  z <- pars[2]
 
  # add your code here, I set up the params for you this time
  
  predicted_N <- rep(NA, length(dat$year))
  
  predicted_N[1] <- K
  
  for (i in 2:length(dat$year)) {
    predicted_N[i] <- predicted_N[i-1] + r * predicted_N[i-1] * (1 - (predicted_N[i-1] / K)^z) - dat$catches[i-1]
  }
  
  return(predicted_N)
}

# How well do these initial parameter values perform?
# Also, a way to test that our function is working correctly

mod5_predN <- model5(pars,gray)

# add your plot code here

plot(x=gray$year, 
     y=mod5_predN,
     type="l",
     ylim=c(0,100000),
     xlab="Year",
     ylab="Abundance",
     lwd=2)

points(x=gray$year,
       y=gray$estimated_abundance,
       pch=19)

#### Statistical Model 5 Code - Estimating best parameter estimates ####
# Written for your review, no need to edit it for this assignment



# The statistical model! 
fn_opt <- function(pars,tdat=gray) { # short hand for "function to optimize"
  
  pred <- model5(pars, tdat)
  nll <- sum(log((pred/tdat[,2]))^2, na.rm=T)
  
  z <- exp(pars[2])
  if(z >35 | z <1) nll <- Inf
  return(nll)
}

model_5 <- optim(pars, fn_opt)

# The best estimates for the three parameters
best.param5 <- model_5$par
exp(best.param5)

# Plot how well these predictions fit to the data when using best param estimates
mod5best_predN <- model5(best.param5, gray)
lines(x=gray$year, y=mod5best_predN, lwd = 2, col="red")

```


# Model 6: Pella-Tomlinson with inflated catches

```{r}

pars <- log(c(.01, 1, 10)) # initial param values

model6 <- function(pars, dat) {
  pars <- exp(pars)
  
  r <- pars[1]
  K <- 100670
  z <- pars[2]
  X <- pars[3]
   
  predicted_N <- rep(NA, length(dat$year))
  
  predicted_N[1] <- K
  
    for (i in 2:length(dat$year)) {
    if(dat$year[i] <= 1950){
   predicted_N[i] <- predicted_N[i-1] + r * predicted_N[i-1] * (1 - (predicted_N[i-1] / K)^z) - (dat$catches[i-1]*X)
    }
    else{
   predicted_N[i] <- predicted_N[i-1] + r * predicted_N[i-1] * (1 - (predicted_N[i-1] / K)^z) - dat$catches[i-1]
    }
      if (predicted_N[i] <= 0) {
      predicted_N[i] <- 0
    }

 }
  
  return(predicted_N)
}


# How well do these initial parameter values perform?
# Also, a way to test that our function is working correctly

mod6_predN <- model6(pars,gray)


# add your plot code here
plot(x=gray$year, 
     y=mod6_predN,
     type="l",
     ylim=c(0,100000),
     xlab="Year",
     ylab="Abundance",
     lwd=2)

points(x=gray$year,
       y=gray$estimated_abundance,
       pch=19)


#### Statistical Model 6 Code - Estimating best parameter estimates ####
# Written for your review, no need to edit it for this assignment


# The statistical model! 
fn_opt <- function(pars,tdat=gray) {
  
  pred <- model6(pars, tdat)
  nll <- sum(log((pred/tdat[,2]))^2,na.rm=T)
  
  z <- exp(pars[2])
  if(z >35 | z <1) nll <- Inf
  return(nll)
}

model_6 <- optim(pars, fn_opt)

# The best estimates for the three parameters
best.param6 <- model_6$par
exp(best.param6)

# Plot how well these predictions fit to the data when using best param estimates
mod6best_predN <- model6(best.param6, gray)
lines(x=gray$year, y=mod6best_predN, lwd=2, col="red")
```

## Part 3: How do the model fits look?

According to our graphs, model 6, Pella-Tomlinson with inflated catches fit the model best. This was followed by Model 5: Pella-Tomlinson 0.5-0.9K with N0=K, which diverged slightly through time from it's coded model. It would be interesting to analyze these predicted models with their coded models using a sensitivity analysis to better understand how they differ (the predicted models from the coded models).

